Sequential Composition for Relaxed Memory: Pomsets with Predicate Transformers

### 2021-04-22 paper/arm.tex
added second arm example in appendix

  Execution $\aEG$ is
  \emph{$({\rco}, {\rrfx}, {\rcb})$-valid} under \emph{External Consistency} (\EC{}) if
  \begin{enumerate}[resume,label=(\textsc{a}\arabic*),ref=\textsc{a}\arabic*]
  \item[\eqref{arm-co}] and \eqref{arm-rf}, as for \EGC,
  \item \label{arm-cb}
    ${\rcb}\supseteq\PBR{{\rco}\cup{\rlob}}$ is a linear order %, capturing \emph{completes before}, %$\rcb : % \Event\times\Event$
    such that if $\bEv\xrfx\aEv$ then either
    \begin{enumerate}%[leftmargin=0pt]
    \item \label{arm-rfe}
      % $\bEv\xcb\aEv$ and $\cEv\rblocks\aEv$ then either $\cEv\xcb\bEv$ or $\aEv\xcb\cEv$, or 
      $\bEv\xcb\aEv$ and if $\labeling(\cEv) \rblocks \labeling(\aEv)$ then either $\cEv\xcb\bEv$ or $\aEv\xcb\cEv$, or
      % $\bEv\xcb\aEv$ and $(\not\exists\cEv)$ such that $\bEv\xcb\cEv\xcb\aEv$ and $\cEv\rblocks\aEv$, or 
    \item \label{arm-rfi}
      $\bEv\xcbinv\aEv$ and $\bEv\xpoloc\aEv$ and $(\not\exists\cEv)$ $\labeling(\cEv) \rblocks \labeling(\aEv)$ and $\bEv\xpoloc\cEv\xpoloc\aEv$.
      % $\bEv\xpoloc\aEv$ and $(\not\exists\cEv)$ such that $\bEv\xpoloc\cEv\xpoloc\aEv$ and $\cEv\rblocks\aEv$.
      % \item if $\bEv\xco\aEv$ then $\bEv\xgcb\aEv$,
      % \item if $\bEv\xrfx\aEv$ then $\bEv\xgcb\aEv$,
      % \item if $\bEv\xrfx\aEv$ and $\cEv\rblocks\aEv$ then either $\cEv\xgcb\bEv$ or $\aEv\xgcb\cEv$,
      % \item if $\bEv\xlob\aEv$ then either $\bEv\xgcb\aEv$ or $(\exists\cEv)$
      %   $\cEv\xrfx\bEv$ and $\cEv\xpoloc\bEv$ but not $\cEv\xlob\aEv$.
    \end{enumerate}
  \end{enumerate}
\end{definition}
\citet{armed} show that \EGC{} and \EC{} are both equivalent to the standard
definition of Arm8.  They explain \EGC{} and \EC{} using the following
example, which is allowed by Arm8.\footnote{We have changed an address
  dependency in the first thread to a data dependency.}
\begin{gather*}
  \PW{x}{1}\SEMI 
  \PR{x}{r}\SEMI
  \PW{y}{r} \PAR
  \PR[\mACQ]{y}{1}\SEMI
  \PR{x}{s}
  \\
  %\tag{\cmark\armeight}
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{a}{\DW{x}{1}}{}
      \event{b}{\DR{x}{1}}{right=of a}
      \event{c}{\DW{y}{1}}{right=of b}
      \raevent{d}{\DR[\mACQ]{y}{1}}{right=2.5em of c}
      \event{e}{\DR{x}{0}}{right=of d}
      \rfx{a}{b}
      \lob{b}{c}
      \rfx{c}{d}
      \lob{d}{e}
      \co[out=-165,in=-15]{e}[above,pos=.375]{a}
    \end{tikzinline}}
\end{gather*}
\EGC{} drops $\rlob$-order in the first thread using \ref{arm-gcb-lob}, since
$\DWP{x}{1}$ is not $\rlob$-ordered before $\DWP{y}{1}$.
\begin{gather*}
  \tag{$\rgcb$}
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{a}{\DW{x}{1}}{}
      \event{b}{\DR{x}{1}}{right=of a}
      \event{c}{\DW{y}{1}}{right=of b}
      \raevent{d}{\DR[\mACQ]{y}{1}}{right=2.5em of c}
      \event{e}{\DR{x}{0}}{right=of d}
      \gcbz{a}{b}
      \gcbz{c}{d}
      \gcbz{d}{e}
      \gcbz[out=-165,in=-15]{e}{a}
      % \rfx{a}{b}
      % %\lob{b}{c}
      % \rfx{c}{d}
      % \lob{d}{e}
      % \co[out=-165,in=-15]{e}[above right]{a}
    \end{tikzinline}}
\end{gather*}
\EC{} drops $\rrfx$-order in the first thread using \ref{arm-rfi}.
\begin{gather*}
  \tag{$\rcb$}
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{a}{\DW{x}{1}}{}
      \event{b}{\DR{x}{1}}{right=of a}
      \event{c}{\DW{y}{1}}{right=of b}
      \raevent{d}{\DR[\mACQ]{y}{1}}{right=2.5em of c}
      \event{e}{\DR{x}{0}}{right=of d}
      \cbz{b}{c}
      \cbz{c}{d}
      \cbz{d}{e}
      \cbz[out=-165,in=-15]{e}{a}
      % %\rfx{a}{b}
      % \lob{b}{c}
      % \rfx{c}{d}
      % \lob{d}{e}
      % \co[out=-165,in=-15]{e}[above right]{a}
    \end{tikzinline}}
\end{gather*}

@@

If the read of $x$ in the execution above is changed from acquiring to
relaxed, then our semantics allows the execution, using the independent case
for the read and satisfying the precondition of $\DWP{y}{1}$ by prepending
$\DWP{x}{2}$.  It may be tempting, therefore to adopt a strategy of
\emph{downgrading} acquires in certain cases.  Unfortunately, it is not
possible to do this locally without violating important idioms such as
publication.  For example, consider that $\DRP[\mRA]{x}{1}$ \emph{not} possible for
the second thread in the following attempted execution, due to publication of
$\DWP{x}{2}$ via $y$:
\begin{gather*}
  \PW{x}{\PR{x}{}+1}\SEMI
  \PW[\mREL]{y}{1}
  \PAR
  \PW{x}{1}\SEMI
  \IF{\PR[\mACQ]{y}{}\AND\PR[\mACQ]{x}{}}\THEN
  \PR{z}{s}
  \FI
  \PAR
  \PW{z}{1}\SEMI
  \PW[\mREL]{x}{1}
  \\
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{b1}{\DR{x}{1}}{}
      \event{b2}{\DW{x}{2}}{right=of b1}
      \raevent{b3}{\DW[\mREL]{y}{1}}{right=of b2}
      \po{b1}{b2}
      \sync{b2}{b3}
      \event{a1}{\DW{x}{1}}{right=3em of b3}
      \raevent{a2}{\DR[\mACQ]{y}{1}}{right=of a1}
      \raevent{a3}{\DR[\mACQ]{x}{1}}{right=of a2}
      \event{a4}{\DR{z}{0}}{right=of a3}
      \sync{a2}{a3}
      \sync{a3}{a4}
      \event{c1}{\DW{z}{1}}{right=3em of a4}
      \raevent{c2}{\DW[\mREL]{x}{1}}{right=of c1}
      \sync{c1}{c2}
      \rf[out=-165,in=-15]{a1}{b1}
      \rf[out=15,in=165]{b3}{a2}
      \rf[out=-165,in=-15]{c2}{a3}
      \wk{a4}{c1}
    \end{tikzinline}}
\end{gather*}
Instead, if the read of $x$ is relaxed, then the publication via $y$ fails,
and $\DRP{x}{1}$ in the second thread is possible.
\begin{gather*}
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{b1}{\DR{x}{1}}{}
      \event{b2}{\DW{x}{2}}{right=of b1}
      \raevent{b3}{\DW[\mREL]{y}{1}}{right=of b2}
      \po{b1}{b2}
      \sync{b2}{b3}
      \event{a1}{\DW{x}{1}}{right=3em of b3}
      \raevent{a2}{\DR[\mACQ]{y}{1}}{right=of a1}
      \event{a3}{\DR{x}{1}}{right=of a2}
      % \raevent{a3}{\DR[\mACQ]{x}{1}}{right=of a2}
      \event{a4}{\DR{z}{0}}{right=of a3}
      \sync{a2}{a3}
      \sync[out=15,in=165]{a2}{a4}
      \event{c1}{\DW{z}{1}}{right=3em of a4}
      \raevent{c2}{\DW[\mREL]{x}{1}}{right=of c1}
      \sync{c1}{c2}
      \rf[out=-165,in=-15]{a1}{b1}
      \rf[out=15,in=165]{b3}{a2}
      \rf[out=-165,in=-15]{c2}{a3}
      \wk{a4}{c1}
    \end{tikzinline}}
\end{gather*}

>>>>

  Execution $\aEG$ is
  \emph{$({\rco}, {\rrfx}, {\rcb})$-valid} under \emph{External Consistency} (\EC{}) if
  \begin{enumerate}[resume,label=(\textsc{a}\arabic*),ref=\textsc{a}\arabic*]
  \item[\eqref{arm-co}] and \eqref{arm-rf}, as for \EGC,
  \item \label{arm-cb}
    ${\rcb}\supseteq\PBR{{\rco}\cup{\rlob}}$ is a linear order %, capturing \emph{completes before}, %$\rcb : % \Event\times\Event$
    such that if $\bEv\xrfx\aEv$ then either
    \begin{enumerate}%[leftmargin=0pt]
    \item \label{arm-rfe}
      % $\bEv\xcb\aEv$ and $\cEv\rblocks\aEv$ then either $\cEv\xcb\bEv$ or $\aEv\xcb\cEv$, or 
      $\bEv\xcb\aEv$ and if $\labeling(\cEv) \rblocks \labeling(\aEv)$ then either $\cEv\xcb\bEv$ or $\aEv\xcb\cEv$, or
      % $\bEv\xcb\aEv$ and $(\not\exists\cEv)$ such that $\bEv\xcb\cEv\xcb\aEv$ and $\cEv\rblocks\aEv$, or 
    \item \label{arm-rfi}
      $\bEv\xcbinv\aEv$ and $\bEv\xpoloc\aEv$ and $(\not\exists\cEv)$ $\labeling(\cEv) \rblocks \labeling(\aEv)$ and $\bEv\xpoloc\cEv\xpoloc\aEv$.
      % $\bEv\xpoloc\aEv$ and $(\not\exists\cEv)$ such that $\bEv\xpoloc\cEv\xpoloc\aEv$ and $\cEv\rblocks\aEv$.
      % \item if $\bEv\xco\aEv$ then $\bEv\xgcb\aEv$,
      % \item if $\bEv\xrfx\aEv$ then $\bEv\xgcb\aEv$,
      % \item if $\bEv\xrfx\aEv$ and $\cEv\rblocks\aEv$ then either $\cEv\xgcb\bEv$ or $\aEv\xgcb\cEv$,
      % \item if $\bEv\xlob\aEv$ then either $\bEv\xgcb\aEv$ or $(\exists\cEv)$
      %   $\cEv\xrfx\bEv$ and $\cEv\xpoloc\bEv$ but not $\cEv\xlob\aEv$.
    \end{enumerate}
  \end{enumerate}
\end{definition}
\citet{armed} show that \EGC{} and \EC{} are both equivalent to the standard
definition of Arm8.  They explain \EGC{} and \EC{} using the following
example, which is allowed by Arm8.\footnote{We have changed an address
  dependency in the first thread to a data dependency.}
\begin{gather*}
  \PW{x}{1}\SEMI 
  \PR{x}{r}\SEMI
  \PW{y}{r} \PAR
  \PR[\mACQ]{y}{1}\SEMI
  \PR{x}{s}
  \\
  %\tag{\cmark\armeight}
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{a}{\DW{x}{1}}{}
      \event{b}{\DR{x}{1}}{right=of a}
      \event{c}{\DW{y}{1}}{right=of b}
      \raevent{d}{\DR[\mACQ]{y}{1}}{right=2.5em of c}
      \event{e}{\DR{x}{0}}{right=of d}
      \rfx{a}{b}
      \lob{b}{c}
      \rfx{c}{d}
      \lob{d}{e}
      \fr[out=-165,in=-15]{e}[above,pos=.375]{a}
    \end{tikzinline}}
\end{gather*}
\EGC{} drops $\rlob$-order in the first thread using \ref{arm-gcb-lob}, since
$\DWP{x}{1}$ is not $\rlob$-ordered before $\DWP{y}{1}$.
\begin{gather*}
  \tag{$\rgcb$}
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{a}{\DW{x}{1}}{}
      \event{b}{\DR{x}{1}}{right=of a}
      \event{c}{\DW{y}{1}}{right=of b}
      \raevent{d}{\DR[\mACQ]{y}{1}}{right=2.5em of c}
      \event{e}{\DR{x}{0}}{right=of d}
      \gcbz{a}{b}
      \gcbz{c}{d}
      \gcbz{d}{e}
      \gcbz[out=-165,in=-15]{e}{a}
      % \rfx{a}{b}
      % %\lob{b}{c}
      % \rfx{c}{d}
      % \lob{d}{e}
      % \co[out=-165,in=-15]{e}[above right]{a}
    \end{tikzinline}}
\end{gather*}
\EC{} drops $\rrfx$-order in the first thread using \ref{arm-rfi}.
\begin{gather*}
  \tag{$\rcb$}
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{a}{\DW{x}{1}}{}
      \event{b}{\DR{x}{1}}{right=of a}
      \event{c}{\DW{y}{1}}{right=of b}
      \raevent{d}{\DR[\mACQ]{y}{1}}{right=2.5em of c}
      \event{e}{\DR{x}{0}}{right=of d}
      \cbz{b}{c}
      \cbz{c}{d}
      \cbz{d}{e}
      \cbz[out=-165,in=-15]{e}{a}
      % %\rfx{a}{b}
      % \lob{b}{c}
      % \rfx{c}{d}
      % \lob{d}{e}
      % \co[out=-165,in=-15]{e}[above right]{a}
    \end{tikzinline}}
\end{gather*}

@@

If the read of $x$ in the execution above is changed from acquiring to
relaxed, then our semantics allows the $\rgcb$ execution, using the independent case
for the read and satisfying the precondition of $\DWP{y}{1}$ by prepending
$\DWP{x}{2}$.  It may be tempting, therefore to adopt a strategy of
\emph{downgrading} acquires in certain cases.  Unfortunately, it is not
possible to do this locally without violating important idioms such as
publication.  For example, consider that $\DRP[\mRA]{x}{1}$ \emph{not} possible for
the second thread in the following attempted execution, due to publication of
$\DWP{x}{2}$ via $y$:
\begin{gather*}
  \PW{x}{\PR{x}{}+1}\SEMI
  \PW[\mREL]{y}{1}
  \PAR
  \PW{x}{1}\SEMI
  \IF{\PR[\mACQ]{y}{}\AND\PR[\mACQ]{x}{}}\THEN
  \PR{z}{s}
  \FI
  \PAR
  \PW{z}{1}\SEMI
  \PW[\mREL]{x}{1}
  \\
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{b1}{\DR{x}{1}}{}
      \event{b2}{\DW{x}{2}}{right=of b1}
      \raevent{b3}{\DW[\mREL]{y}{1}}{right=of b2}
      \po{b1}{b2}
      \sync{b2}{b3}
      \event{a1}{\DW{x}{1}}{right=3em of b3}
      \raevent{a2}{\DR[\mACQ]{y}{1}}{right=of a1}
      \raevent{a3}{\DR[\mACQ]{x}{1}}{right=of a2}
      \event{a4}{\DR{z}{0}}{right=of a3}
      \sync{a2}{a3}
      \sync{a3}{a4}
      \event{c1}{\DW{z}{1}}{right=3em of a4}
      \raevent{c2}{\DW[\mREL]{x}{1}}{right=of c1}
      \sync{c1}{c2}
      \rf[out=-165,in=-15]{a1}{b1}
      \rf[out=15,in=165]{b3}{a2}
      \rf[out=-165,in=-15]{c2}{a3}
      \wk{a4}{c1}
    \end{tikzinline}}
\end{gather*}
Instead, if the read of $x$ is relaxed, then the publication via $y$ fails,
and $\DRP{x}{1}$ in the second thread is possible.
\begin{gather*}
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{b1}{\DR{x}{1}}{}
      \event{b2}{\DW{x}{2}}{right=of b1}
      \raevent{b3}{\DW[\mREL]{y}{1}}{right=of b2}
      \po{b1}{b2}
      \sync{b2}{b3}
      \event{a1}{\DW{x}{1}}{right=3em of b3}
      \raevent{a2}{\DR[\mACQ]{y}{1}}{right=of a1}
      \event{a3}{\DR{x}{1}}{right=of a2}
      % \raevent{a3}{\DR[\mACQ]{x}{1}}{right=of a2}
      \event{a4}{\DR{z}{0}}{right=of a3}
      \sync{a2}{a3}
      \sync[out=15,in=165]{a2}{a4}
      \event{c1}{\DW{z}{1}}{right=3em of a4}
      \raevent{c2}{\DW[\mREL]{x}{1}}{right=of c1}
      \sync{c1}{c2}
      \rf[out=-165,in=-15]{a1}{b1}
      \rf[out=15,in=165]{b3}{a2}
      \rf[out=-165,in=-15]{c2}{a3}
      \wk{a4}{c1}
    \end{tikzinline}}
\end{gather*}
